---
title: "PA1_template.Rmd"
author: "owilks"
date: "March 30, 2016"
output: html_document

---

## Retrieving the Data

First step was to retrieve the data from the coursepage at: <https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1> and then download and store the data into a table using the following code.

```{r echo=TRUE}
library(chron)
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","c5_a1.zip")
unzip("c5_a1.zip")

data <- read.csv("activity.csv")

```

##What is the mean total number of steps taken per day

*1. Calculate the total number of steps taken per day*

To do this we use the aggregate function to collapse the data based on a function and a variable. Similar to the "GROUP BY" function in SQL. Applying sum on the entire list based on steps and grouped by date.

```{r echo=TRUE}
his1 <- aggregate(data$steps, by=list(date=data$date), FUN=sum, na.rm=TRUE)
```

*2. Make a histogram of the total number of steps taken each day*

```{r echo=TRUE}
hist(his1$x)
```

*3. Calculate and report the mean and median of the total number of steps taken per day*

The mean number of steps taken per day is `r signif(mean(his1$x),4) `

The median number of steps taken per day is `r signif(median(his1$x),4)`

##What is the average daily activity pattern?

*1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)*

Using the same strategy as before, the aggregation was done instead on the "interval" column and the "steps" column.

The function used instead of "sum" this time was "mean" to effectively calculate the requested mean.

```{r echo=TRUE}
ts1 <- aggregate(data$steps, by=list(interval=data$interval), FUN=mean, na.rm=TRUE)

#Instead of using the ts function we directly use the plot function since the
#given list has the x and y values and the instructions indicated the "l" type

plot(ts1$interval,ts1$x,type="l")
```

*2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?*

The first step was finding out which point in the list had the maximum value for steps, which
was accomplished using the **which.max** function to get a value of `r which.max(ts1$x)`.

The 5-min interval with the maximum number of steps was **ts1$interval[104]** or `r ts1$interval[104]`

##Inputing missing values

*1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)*

Using the **is.na** function we converted our "data" data frame into a data frame of True/False values, values that could be summed to count the number of NA rows.This can also be done by counting the NA values in the individual columns.

```{r echo=TRUE}
tf = is.na(data)
na_steps <- sum(is.na(data$steps))
na_interval <- sum(is.na(data$interval))
na_date <- sum(is.na(data$date))
```

The total number of NA values  are `r na_steps` , `r na_interval` and `r na_date` respective to their order in the code so NA values are only found in the steps column.

*2. Devise a strategy for filling in all of the missing values in the dataset. *
*3. Create a new dataset that is equal to the original dataset but w/o missing data. *

Replacing the NA values with the medians for all the days is sufficient to update the data set.

```{r echo=TRUE}
data_fix <- data
data_fix$steps[is.na(data_fix$steps)] <- mean(data_fix$steps, na.rm = TRUE)
```

*4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.*

```{r echo=TRUE}
his2 <- aggregate(data_fix$steps, by=list(date=data_fix$date), FUN=sum, na.rm=TRUE)
hist(his2$x)
```

The new mean number of steps taken per day is `r signif(mean(his2$x),4) `

The new median number of steps taken per day is `r signif(median(his2$x),4)`

The NA values going toward the mean skews the entire data set towards the center, as visible in the differences between the histograms.

##Are there differences in activity patterns between weekdays and weekends?

*1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day*

First step was calculating whether or not the day was on the weekend using the **"is.weekend"** function and storing that in a new variable using the **ifelse** function.

```{r echo=TRUE}
data_fix$wknd <- ifelse(is.weekend(as.Date(data_fix$date)),"weekend","weekday")
data2 <-data_fix

```

Then we create ts_day and ts_end to create the underlying data framework for creating the time series plots.

To use the factors accordingly will be using the **subset** function to isolate all the weekday data versus all the weekend data for the respective newly created data.frames.

```{r echo=TRUE}

ts_day <- aggregate(data2$steps, by=list(interval=data2$interval), FUN=mean, data = subset(data2,data2$wknd=="weekday"), na.rm=TRUE)

#That additional argument effectively subsets the data based on the wknd values

ts_end <- aggregate(data2$steps, by=list(interval=data2$interval), FUN=mean, data = subset(data2,data2$wknd=="weekend"), na.rm=TRUE)
```

*2. Make a panel plot containing a time series plot (i.e. type = "l")*

Using the **par** function we can create plots much like the previous box plots but allow them to be combined.
```{r echo=TRUE}
par(mfrow=c(2,1))
plot(ts_day$interval,ts_day$x,type="l")
plot(ts_end$interval,ts_end$x,type="l")